Inference for Trans-dimensional Bayesian Models with Diffusive Nested Sampling

### 2014-08-17 Documents/Paper/ms.tex
GalaxyField N figure
The two phases
Paragraph

In other studies, it has been common to do separate runs of Nested Sampling,
with $N$ fixed at various trial values.
Then, the posterior for $N$ can be calculated
based on the estimates of the evidence obtained as a function of $N$. However,
in this paper, we use trans-dimensional MCMC moves to infer $N$. In other words,
$N$ is treated as just another parameter, and the marginal likelihood defined
in Equation~\ref{eq:evidence}, which we can compute, also includes the sum over
the unknown parameter $N$. Our motivation
for using Nested Sampling is not primarily to calculate $\mathcal{Z}$.
The motivation for using Nested Sampling, rather than just sampling the
posterior using trans-dimensional MCMC, is that the posterior often contains
difficult features, such as strong dependencies or multiple modes, that cause
problems with mixing. Diffusive Nested Sampling
(DNS) replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained distributions, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of simulated annealing,
where the target distribution is modified from the posterior to something
``easier'', however,
the sequence of distributions used in Nested Sampling avoids some of the
problems with simulated annealing: in particular, there is no need to choose
a temperature schedule, and the method does not fail when phase transitions
are present.

@@

\section{Metropolis proposals for the general problem}
We will now define a set of proposal distributions that can be used to
sample the prior distribution
$p(N, \hyperparams, \{\xx_i\}) = p(N) p(\hyperparams) \prod_{i=1}^N p(\xx_i | \hyperparams)$.
These proposals will need to satisfy detailed balance in order to be valid
when used inside DNS. When the proposal satisfies detailed balance with respect
to the prior, the DNS algorithm can incorporate hard likelihood constraints by
rejecting any proposal whose likelihood does not satisfy the constraint.
The overall proposal is a mixture of all of the proposals listed here.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{galaxyfield_likelihood.pdf}
\caption{{\bf Top panel: }
The shape of the likelihood function with respect to prior mass for the
GalaxyField data.
{\bf Bottom panel: }The posterior weights of the saved samples. Note the phase
transition at $\log(X) \approx -100$ nats. While this doesn't affect the
posterior distribution (unlike the sinewave example), it would affect calculation
of the marginal likelihood if annealing were used.
\label{fig:sinewaves_likelihood}}
\end{center}
\end{figure}

>>>>

In other studies, it has been common to do separate runs of Nested Sampling
(or another method),
with $N$ fixed at various trial values \citep[e.g.][]{fengji}.
Then, the posterior for $N$ can be calculated
based on the estimates of the evidence obtained as a function of $N$. However,
in this paper, we use trans-dimensional MCMC moves to infer $N$. In other words,
$N$ is treated as just another parameter, and the marginal likelihood defined
in Equation~\ref{eq:evidence}, which we can compute, also includes the sum over
the unknown parameter $N$. Our motivation
for using Nested Sampling is not primarily to calculate $\mathcal{Z}$.
The motivation for using Nested Sampling, rather than just sampling the
posterior using trans-dimensional MCMC, is that the posterior often contains
difficult features, such as strong dependencies or multiple modes, that cause
problems with mixing. Diffusive Nested Sampling
(DNS) replaces the posterior distribution with an alternative target
distribution composed of a mixture of the prior distribution with other, more
constrained distributions, facilitating mixing between the multiple modes
and along degeneracy curves.
This is similar to the idea of simulated annealing,
where the target distribution is modified from the posterior to something
``easier'', however,
the sequence of distributions used in Nested Sampling avoids some of the
problems with simulated annealing: in particular, there is no need to choose
a temperature schedule, and the method does not fail when phase transitions
are present.

Since DNS is effectively the Metropolis-Hastings algorithm applied to a
distribution other than the posterior, proposal distributions are needed.
In Section~\ref{sec:proposals}, we outline a set of proposal distributions used
to explore the hypothesis space of possible values for $N$, $\hyperparams$,
and $\{\xx_i\}$.

@@

\section{Metropolis proposals for the general problem}\label{sec:proposals}
We will now define a set of proposal distributions that can be used to
sample the prior distribution
$p(N, \hyperparams, \{\xx_i\}) = p(N) p(\hyperparams) \prod_{i=1}^N p(\xx_i | \hyperparams)$.
These proposals will need to satisfy detailed balance in order to be valid
when used inside DNS. When the proposal satisfies detailed balance with respect
to the prior, the DNS algorithm can incorporate hard likelihood constraints by
rejecting any proposal whose likelihood does not satisfy the constraint.
The overall proposal is a mixture of all of the proposals listed here.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{galaxyfield_likelihood.pdf}
\caption{{\bf Top panel: }
The shape of the likelihood function with respect to prior mass for the
GalaxyField data. Note the phase
transition at $\log(X) \approx -100$ nats. This separates models with just
the few brightest galaxies from models with many faint galaxies as well.
While this doesn't affect the
posterior distribution (unlike the sinewave example), it would affect calculation
of the marginal likelihood if annealing were used.
{\bf Bottom panel: }The posterior weights of the saved samples.
\label{fig:sinewaves_likelihood}}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{N_galaxy_result.pdf}
\caption{The inference for $N$, the number of galaxies, based on the
GalaxyField data. The true value was $N=47$.
\label{fig:N_galaxy_result}}
\end{center}
\end{figure}
